Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-17 draft.tex
added Bernhards correction and improved Abstrackt with Steffens feedback

%% Abstract section.
\abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{interactions}).
%
Modern cohort studies also incorporate medical image data.
%
Analyzing these data requires image segmentation, extraction of key figures and shape based subject grouping.
\\\\
We propose a Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as sociodemographic and medical attribute data.
%
It allows for both hypothesis validation approaches as well as hypothesis generation by incorporating data mining methods.
%
Adaptive linked information visualization views and 3d-shape renderings are combined with epidemiological techniques.
%
Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
%
Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
} % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims to characterize health and disease by determining risk factors.
%
Clinical problems and questions answered using epidemiological methods comprise diagnosis accuracy, disease frequency, risk factors, disease prognosis, effectiveness of treatments or preventions and cause of diseases \cite{Fletcher2012}.
%
%Cohort studies are a epidemiological method to gather data about a group of subjects (a \emph{cohort)}) to make general statements about these problems.
%
Observations made by clinicians in the daily routine are translated into hypothesis.
%
These are used to determine environmental and lifestyle factors are as well as medical attributes which may influence a condition of interest.
%
The data variables necessary are gathered using structured interviews and clinical examinations.
%
Statistical methods like regression analysis aim aim to check the attribute list for plausibility.
%

Longitudinal population-based studies like the Study of Health in Pomerania \cite{Volzke2011} aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The sample is drawn randomized to avoid selection bias which prohibits statements based on statistical correlations in the cohort from being inferred to the whole population (\emph{external validity}) \cite{Fletcher2012}.
%
Also a information bias needs to be avoided by strictly standardizing the data acquisition.
%
Statistical correlations are also prone to \emph{confounding}, meaning that two factors are dependent and therefore should be normalized with respect to each other.
%
When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
%
Therefore results need to be age adjusted to be comparable.
%
Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
\\\\
Modern cohort studies often include medical image data which introduces new problems.
%
Since it is unethical to expose people to radiation, non-harming imaging like Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
%
As MRI scans are expensive there exists a tradeoff between quality of the image data and their associated costs.
%
To quantify these data it is necessary to label each voxel regarding structure affiliation (\emph{segmentation}).
%
Manual segmentation carried out by radiological experts is possible but very costly and prone to inter- and intra observer variability.
%
Segmentation algorithms allow for (semi)-automated analysis of the data but require sophisticated methods due to high inter-subject variability caused by the subject diversity.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.
\\\\
We propose a Interactive Visual Analysis approach \cite{Thomas2005} to provide a way to analyze both image- and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space.
%
Intended as a extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger hypothesis generation using Data Mining methods such as clustering.
%
In order to characterize the healthy aging process we aim to determine changes for subjects which indicate unusual pathological changes.
\\\\
Our contributions are:
\begin{itemize}
	\item Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing affordances of this context.
	\item Provide an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
	\item Provide visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.
	\item Implement the presented methods as a web framework based on WebGL, D3JS and NodeJS.
\end{itemize}

@@

In this section we want to give insight into the epidemiological workflow when analyzing cohort study data to identify the problems we address in this paper.
%
% TODO Define Epidemiological Outcome - 5Ds:
% Death
% Disease
% Discomfort
% Disability
% Dissatisfaction
%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
% TODO Define who participates in epidemiological studies ?
Since it is a diverse science, many different experts work at epidemiological studies, ranging from specialized doctors, medical computer scientists with focus on biometrics to statisticians.
%
Epidemiologists follow a strict workflow mainly driven by statistic tools to validate hypothesis about disease specific risk factors.
%
Following Thew and colleagues publication on this matter, the workflow can be characterized as follows.
%
\begin{itemize}
	\item Hypothesis most commonly base on observations made by clinicians in their daily routine.
%
	\item A set of attributes depicting conditions affected by the hypothesis is compiled accordingly.
%
	\item Confounding variables need to be adjusted so that they do not affect the effect size of a attribute.
%
	\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
\end{itemize}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a).

Reproducibility of results is an key requirement.
%
Longitudinal studies require the acquired attributes to be comparable to evaluate them.
%
If the data acquisition process changes, a information bias is introduced to the data, disallowing inference between acquisition cycles.
%
This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.
%
To determine, whether a subject is prone to be affected by a certain disease, relative risks are expressed through the evaluation of p-values which indicate statistical significance.
%
Statistics tools such as SPSS and STATA play a major role for analyzing epidemiological data. 
%
Graphic data representation is largely used to present results rather than gaining insight.

@@

\subsection{Epidemiological Data} \label{EpidemiologicalData}
Epidemiological data is highly heterogenous.
%
Information about medical history and examinations, genetic conditions, geographical data, questionnaire results and image data yields a complex data space for each subject.
%
Often data are derived from acquired variables to either group or threshold values or get information derived from reviewed data such as breast density data for women.
%
This underlines also the problem of missing data since for ethical, legal or medical reasons some data variables can not be gathered for each subject.
%
Follow-up examinations or -questions for conditions also produce variables only available for a small amount of subjects.
%

Indicators for medical conditions as well as questions about a subjects lifestyle are also often \emph{dichotomous}, meaning that they only have two manifestations (often \emph{Yes} or \emph{No}).
%
This allows for the calculation of \emph{odds ratios} which describe the relation of two \emph{dichotomous} variables, allowing for direct comparison of their influences.
%
Dichotomous data can also be derived by combining aggregating data variables to yield only two manifestations (e.g. subjects younger or older than 50).
%

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques emitting an hazardous amount of radiation for the subject are not suited for ethical reasons.
%
MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.

\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentations masks labeling the voxel of a shape of interest would be ideal since many different key figures can be derived from them.
%
Since these masks require sophisticated algorithms custom tailored to the data sets the epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
%
Information derived by landmarks are also not nearly as expressive and versatile as segmentation masks.
%
They are also prone to a high inter-observer variability and hard to reproduce.
%
This gains even more momentum when analyzing multiple time steps!
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determine type of tissue).

@@

\subsection{The Study of Health in Pomerania (SHIP)}
Starting 1997 with a cohort consisting of 4.308 subjects this cohort study located in northern Germany aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data is collected diseases-independent.
%
This allows the data set to be queried regarding many different diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration of 2012.
%
The MRI protocol features a rich number of different sequences.
%
Also for women, breast MRI scans were acquired.
%
A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
%
The protocols for analyzing the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
%
The overall examination time for each person attending the study is two days.

@@

Designing a visualization which communicates all aspects of the data equally is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, it is often a good solution to combine the strength of different visualization techniques in a unified system \cite{Buja91, Konyha2009}.
%
Data mining tools like the Principal Component Analysis are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to determine.
\\\\
The work of Turkay and colleagues is closest to ours albeit our focus on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
%
Investigating Data on an norwegian aging study their methods aim to amplify a hypothesis generation process.
%
Statistical measures of metric variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots}.
%
These transform dimensions into data points and make them comparable with respect to the derived descriptive measures.
%
This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
The technique was applied to variables generated by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.
%
Hypothesis based on observations of changes in these plots may impose over-fitting to the data because the measure highlights only subsets of statistical changes.
%
Our approach sticks more to the information extracted from the segmented image data and derive variable interaction with non-image epidemiological factors.
% Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!

Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems which analyzed medical image and non-image data using linked views \cite{Gresh2000}.
%
Blaas and colleagues presented a similar system which analyzed medical image data and variables derived from them using views from the feature- and physical space \cite{Blaas2007}.
%
This approach already incorporated Data Mining approaches like dividing the data space using a k-nearest-neighbor technique and Principal Component Analysis.
%
Steenwijk and colleagues employ a relational database to organize the data to visualize subject data using linked views like parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
%
Zhang and colleagues provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set \cite{Zhang2012}.
%
Their understanding of a cohort differs from the understanding of the term in an epidemiolgical context.

Commercial systems like \texttt{Tableau} or \texttt{Spotfire} provide a rich user interface that allow to apply Visual Analytics techniques without the need of writing any code.
%
With little effort, linked views can be created using these tools, but the data processing possibilities like derivation of new variables or the volume rendering capabilities are very limited.
%
These systems share limitations in extensibility to a specific problem domain.
%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language

% *********************** Our own Work ***********************
Klemm and colleagues used lumbar spine variabilities based on an semi-automatic shape-detection algorithm of 490 participants of the \texttt{SHIP-2} \cite{Klemm2013VMV}.
%
Hierarchical agglomerative clustering divided the population into shape-related groups.
%
As proof of concept a relation between size of the segmented shape and measured size of the subjects was measured and behaved as expected.
%
This work focuses on incorporating these data as new features of the overall data set, making it possible to include it into the hypothesis validation and generation process.
%
When applying clustering techniques on the non-image data is was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
% *********************** / Our own Work ***********************

Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique that allows for heterogenous variables to be pairwise compared using appropriate plots in a plot matrix grouped by type \cite{Francois2013}.
%
This technique is also useful to gain an overview over numerous variables and their distributions.
%
It uses histograms, bar charts, scatterplots and heat maps to visualize the different variable combinations with regard to their type.
%
Brushing capabilities allow for brushing and linking as well as filtering, but has limitations like making only one category brushable at a time.
%
We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variable but does not fit in the scope of this paper.
%
The inspiration on the chosen visualization techniques stems from this publication.
%
A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g. mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
%
From their findings regarding the interaction of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.

Chui and colleagues visualized interactions in time-dependant epidemiological data using time-series plots highlighting risk factors differences in age and gender \cite{Chui2011}.
%
\\\\
% Image Data
Comparing tissue between many subjects in an epidemiological context requires methods which allow for shape variance visualizations.
%
Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded in their user study that user favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
%
The distribution of shapes in a space derived from a principal component analysis is plotted by Busking and Colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
%
We incorporate the idea of combining 3D-Shape rendering with information visualization techniques.
%
Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
%
Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in contour-view.
%
Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
%
The data needs to abstracted or summarized in order to work in this context.
%
In order to detect local deformation changes, Hermann and colleagues investigating shape related difference by letting the user specify a deformation of interest and showing corresponding changes in the shape using covariance tensors \cite{Hermann2014}.
%
This method allowed for rapid hypothesis validation and was able to reproduce textbook knowledge.
%
By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
%
This requires a geographic colocation of associated features.
%

% Interactive Visual Analysis - Steffens Paper
The strength of the Interactive Visual Analysis approach described in the next section is it's versatility with respect to the application field \cite{Konyha2009}.
%
Oeltze and colleagues combined a linked view representation of results from a statistical analysis with feature localizations of the human blood flow with the goal of its evaluation \cite{Oeltze2007}.
%
%They aim to provide methods to analyze perfusion data with respect to 
While we take similar steps when analyzing the data like employing statistical tests, our data is mostly independent from the medical image data and is not describing it--except the variables derived the segmentation model itself.

>>>>

%% Abstract section.
\abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
%Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{interactions}).
Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{interactions}).
%
Modern cohort studies also incorporate medical image data.
%
Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
%
The statistically driven epidemiological workflow only allows to determine \emph{interactions} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
\\\\
We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
%
Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.
%As extension to the epidemiological workflow it aims to unveil interactions based on all reliable shape information rather than abstracting it to a few key figures.
%Iterative analysis between ... allows for extracting interactions which without the visiual analysis driven approach maybe would be overlooked. %Einsicht in Interaktionen zwischen nicht-bilddaten und bilddaten auf basis aller verfügbarer Informationen 
%Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
%Iterative analysis of image-based epidemiological datasets aims to provide insight into interactions between organ shape and % in order to provide insights into \emph{interactions} by incorporating additional information
%It extends the epidemiological workflow with an iterative Analysis process and 
It allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
%
Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
%
Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
} % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims to characterize health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond, are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%Clinical problems and questions answered using epidemiological methods comprise diagnosis accuracy, disease frequency, risk factors, disease prognosis, effectiveness of treatments or preventions and cause of diseases \cite{Fletcher2012}.
%
On the other hand, observations made by clinicians in the daily routine are translated into hypothesis for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical attributes which may influence a condition of interest.
%
The data variables necessary are gathered using structured interviews and clinical examinations.
%
Statistical methods like regression analysis aim to check the attribute list for plausibility.
%

Longitudinal population-based studies, such as the Study of Health in Pomerania \cite{Volzke2011}, aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The sample is drawn randomized to avoid selection \cite{Fletcher2012}. %bias which prohibits statements based on statistical correlations in the cohort from being inferred to the whole population (\emph{external validity}) \cite{Fletcher2012}.
%
%Also a information bias needs to be avoided by strictly standardizing the data acquisition.
%
When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
%
Therefore results need to be age-adjusted to be comparable.
%
Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
\\\\
Modern cohort studies often include medical image data.
%
Since it is unethical to expose people to radiation, non-harming imaging such as Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
%
%As MRI scans are time-consuming and expensive there exists a tradeoff between quality of the image data and their associated costs.
%
To quantify these data it is necessary to label each voxel regarding structure affiliation (\emph{segmentation}).
%
Manual segmentation carried out by radiological experts is possible but very costly and prone to inter- and intra observer variability.
%
Segmentation algorithms allow for (semi)-automated analysis of the data but require sophisticated methods due to high inter-subject variability caused by the subject diversity.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.
\\\\
We propose an Interactive Visual Analysis approach \cite{Thomas2005} to provide a way to analyze both image- and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space.
%
Intended as an extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger \emph{hypothesis generation} using Data Mining methods such as clustering.
%
In order to characterize the healthy aging process we aim to determine changes for subjects which indicate unusual pathological changes.
\\\\
Our contributions are:
\begin{itemize}
	\item Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing requirements of this context.
	\item Provide an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
	\item Provide visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.
	\item Implement the presented methods as a web framework based on WebGL, D3JS and NodeJS.
\end{itemize}

@@

In this section we want to give insight into the epidemiological workflow when analyzing cohort study data to identify the problems we address in this paper.
%
% TODO Define Epidemiological Outcome - 5Ds:
% Death
% Disease
% Discomfort
% Disability
% Dissatisfaction
%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypothesis about disease specific risk factors.
%
Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}.
%
\begin{itemize}
	\item Hypothesis most commonly base on observations made by clinicians in their daily routine.
%
	\item A set of attributes depicting conditions affected by the hypothesis is compiled accordingly.
% TODO: Bernhard: Effect size of Attribute ?
	\item Confounding variables need to be adjusted so that they do not affect the effect size of a attribute.
%
	\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
\end{itemize}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a).

Reproducibility of results is an key requirement in epidemiology.
%
Longitudinal studies require the acquired attributes to be comparable to evaluate them.
%
If the data acquisition process changes, a information bias is introduced to the data, hampering inference in acquisition cycles.
%
%This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.
%
To determine, whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that two factors are dependent and therefore should be normalized with respect to each other.
%
Statistics tools such as SPSS and STATA play a major role for analyzing epidemiological data.
%
Graphic data representation is largely used to present results rather than gaining insight.

@@

\subsection{Epidemiological Data} \label{EpidemiologicalData}
Epidemiological data is highly heterogenous and incomplete.
%
Information about medical history and examinations, genetic conditions, geographical data, questionnaire results and image data yields a complex data space for each subject.
%
Often data are derived from acquired variables to either group or threshold values or get information derived from reviewed data such as breast density data for women.
%
This underlines also the problem of missing data since for ethical, legal or medical reasons some data variables can not be gathered for each subject.
%
Follow-up examinations or -questions for conditions also produce variables only available for a small amount of subjects.
%

Indicators for medical conditions as well as questions about a subjects' lifestyle are also often \emph{dichotomous}--they have two manifestations (often \emph{Yes} or \emph{No}).
%
This allows for the calculation of \emph{odds ratios} which describe the relation of two \emph{dichotomous} variables, allowing for direct comparison of their influences.
%
Dichotomous data can also be derived by combining aggregating data variables to yield only two manifestations (e.g. subjects younger or older than 50).
%

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques emitting hazardous amounts of radiation for the subject are not suited for ethical reasons.
%
MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.

\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentation masks labeling the voxel of a shape of interest would be ideal since many different key figures can be derived from them.
%
Since these masks require sophisticated algorithms custom tailored to the data sets the epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
%
Information derived by landmarks are also not nearly as expressive and versatile as segmentation masks.
%
They are also prone to a high inter-observer variability and hard to reproduce.
%
This gains even more momentum when analyzing multiple time steps!
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determine type of tissue).

@@

\subsection{The Study of Health in Pomerania (SHIP)}
Starting 1997 with a cohort consisting of 4.308 subjects this cohort study located in northern Germany aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data is collected without focus on a specific disease.
%
This allows the data set to be queried regarding many different diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration of 2012.
%
The MRI protocol features a rich number of different sequences.
%
Also for women, breast MRI scans were acquired.
%
A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
%
The protocols for analyzing the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
%
The overall examination time for each person attending the study is two days.

@@

Designing a visualization which communicates all aspects of the data equally is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, it is often a good solution to combine the strength of different visualization techniques in a unified system \cite{Buja91, Konyha2009}.
%
The Principal Component Analysis (PCA) and similar techniques are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to determine.
\\\\
The work of Turkay and colleagues is closest to ours albeit our focus on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
%
Investigating Data on an norwegian aging study their methods aim to amplify a hypothesis generation process.
%
Statistical measures of metric variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots}.
%
These transform dimensions into data points and make them comparable with respect to the derived descriptive measures.
%
This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
The technique was applied to variables generated by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.
%
Hypothesis based on observations of changes in these plots may impose over-fitting to the data because the measure highlights only subsets of statistical changes.
%
Our approach sticks more to the information extracted from the segmented image data and derive variable interaction with non-image epidemiological factors.
% Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!

Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems which analyzed medical image and non-image data using linked views \cite{Gresh2000}.
%
Blaas and colleagues presented a similar system which analyzed medical image data and variables derived from them using views from the feature- and physical space \cite{Blaas2007}.
%
This approach already incorporated Data Mining approaches such as dividing the data space using a k-nearest-neighbor technique and Principal Component Analysis.
%
Steenwijk and colleagues employ a relational database to organize the data to visualize subject data using linked views such as parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
%
Zhang and colleagues provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set \cite{Zhang2012}.
%
Their understanding of a cohort differs from the understanding of the term in an epidemiolgical context.

Commercial systems such as \texttt{Tableau} or \texttt{Spotfire} provide a rich user interface that allows to apply Visual Analytics techniques without the need of writing any code.
%
With little effort, linked views can be created using these tools, but the data processing possibilities such as derivation of new variables or the volume rendering capabilities are very limited.
%
These systems share limitations in extensibility to a specific problem domain.
%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language

% *********************** Our own Work ***********************
Klemm and colleagues used lumbar spine variabilities based on an semi-automatic shape-detection algorithm of 490 participants of the \texttt{SHIP-2} \cite{Klemm2013VMV}.
%
Hierarchical agglomerative clustering divided the population into shape-related groups.
%
As proof of concept a relation between size of the segmented shape and measured size of the subjects was measured.
%
This work focuses on incorporating these derived data as new features of the overall data set, making it possible to include it into the hypothesis validation and generation process.
%
When applying clustering techniques on the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
% *********************** / Our own Work ***********************

Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique that allows for heterogenous variables to be pairwise compared using appropriate plots in a plot matrix grouped by type \cite{Francois2013}.
%
This technique is also useful to gain an overview over numerous variables and their distributions.
%
It uses histograms, bar charts, scatterplots and heat maps to visualize the different variable combinations with regard to their type.
%
Brushing, linking and filtering are feasible with \emph{GPLOMS}, but have limitations such as making only one category brushable at a time.
%
%We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variables. %but does not fit in the scope of this paper.
%
%The inspiration on the chosen visualization techniques stems from this publication.
%
A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g. mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
%
From their findings regarding the interaction of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.

Chui and colleagues visualized interactions in time-dependant epidemiological data using time-series plots highlighting risk factors differences in age and gender \cite{Chui2011}.
%
\\\\
% Image Data
Comparing tissue between many subjects in an epidemiological context requires methods which allow for shape variance visualizations.
%
Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded in their user study that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
%
The distribution of shapes in a space derived from a PCA is plotted by Busking and Colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
%
We incorporate the idea of combining 3D-Shape rendering with information visualization techniques.
%
Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
%
Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
%
The data needs to abstracted or summarized in order to work in this context.
%
In order to detect local deformation changes, Hermann and colleagues investigating shape related difference by letting the user specify a deformation of interest and showing corresponding changes in the shape using covariance tensors \cite{Hermann2014}.
%
This method allowed for rapid hypothesis validation and was able to reproduce textbook knowledge.
%
By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
%
This requires a geographic colocation of associated features.
%

% Interactive Visual Analysis - Steffens Paper
The strength of the Interactive Visual Analysis approach described in the next section is its versatility with respect to the application field \cite{Konyha2009}.
%
Oeltze and colleagues combined a linked view representation of results from a statistical analysis with feature localizations of the human blood flow with the goal of its evaluation \cite{Oeltze2007}.
%
%They aim to provide methods to analyze perfusion data with respect to 
While we take similar steps when analyzing the data such as employing statistical tests, our data is mostly independent from the medical image data and is not describing it--except the variables derived the segmentation model itself.

\section{Interactive Visual Analytics in Cohort Study Data}
\begin{figure}[htb]
 \centering
 \label{fig:WorkflowComparison}
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{ToDo Abbildung noch nicht final. Visual Analytics systems are able to complement parts of the epidemiological workflow, not replace it. The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the IVA supported one. The iterative red highlighted part is called the \emph{IVA Loop}.}
\end{figure}
%
As described in subsection~\ref{EpidemiologicalWorkflow}, the epidemiolgical workflow is a sequence of steps taken by domain experts and needs to be reproducible and comprise statistical integrity.
%
Figure~\ref{fig:WorkflowComparison} (a) describes this workflow as consecutive series of steps.
%
The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim to replace the existing workflow but to complement its weaknesses.
%
In the current state the workflow treats the data like a black box.
%
A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
%
The resulting value decides whether the data supports the hypothesis or not.
%
It would be possible that there are actually features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
%
This becomes even more important when the workflow is adapted to the analysis of the medical image data.
%
Domain experts would have to annotate landmarks tediously to derive metrics such as distances. %which are then handled like other features and analyzed using the same set of statistical tools.
%
Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
%
Statistical tests check for validity of the number but not for their completeness or plausibility!
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of the feature list selection process.
%
Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
%
Note that it also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
%
This has to be handled with care since overfitting of expectations to the available data is an imminent danger as described by Turkay and colleagues \cite{Turkay2013}.
